---
title: "Leaflet and OJS: notes and queries"
date: "2026-03-24"
categories: 
  - ojs
  - maps
execute: 
  warning: false
  message: false
  echo: false
---

## Notes

1. an example of a Leaflet map in OJS, using filters and the markerclusters plugin.
2. testing OpenStreetMap basemap to see if I need to change any leaflet settings to avoid Blocked warnings.
3. list of resources


```{r stuff}
library(scales)  
library(janitor) 
library(readxl)
library(glue)

library(tidytext)
library(tidyverse)

ll_url_base <- "https://www.londonlives.org/record/"
```

```{r}

## files in site_data

# improved geo data. drop any existing geo data in cdrv and replace with this.

cdrv_geo_csv <-
  read_csv(here::here("data/ll/cdrv_geocoded_points_and_notgeocoded_20260226.csv"))|>
# add "County " prefix to Irish county names
  mutate(to_county = case_when(
    country=="Ireland" & !is.na(to_county) ~ paste("County", to_county),
    .default = to_county
  ))

# ensure readr treats doc_date as char as there are dates that the parser won't like in there.

cdrv_names_tsv <-
  read_tsv(here::here("data/ll/removal_orders/cdrv_removal_orders_v2_names.tsv"), guess_max = 3100,
           col_types = list(doc_date = col_character(), .default = col_guess()) )

cdrv_orders_tsv <-
  read_tsv(here::here("data/ll/removal_orders/cdrv_removal_orders_v2_orders.tsv"), guess_max = 2500, 
           col_types = list(doc_date = col_character(), .default = col_guess()))


```

```{r}
# get names to add as a list to the orders
cdrv_names <-
cdrv_names_tsv |>
  separate(doc_date, into = c("d1", "d2", "d3"), sep = "-", remove = F) |>
  mutate(date_precision = case_when(
    is.na(doc_date) | doc_date=="" ~ NA,
    d2=="00" ~ "y",
    d3=="00" ~ "ym",
    .default = "ymd"
  )) |>
  # \\b here should cover years like 1700 if needed; but check before using
  mutate(doc_date = str_replace_all(doc_date, "\\b00", "01")) |>
  # make new parsed date col so doc_date remains char.
  mutate(date = parse_date_time(doc_date, "ymd")) |>
  mutate(doc_year = year(date)) |>
  relocate(date, date_precision, doc_year, .after = doc_date) |>
  select(-d1, -d2, -d3)



cdrv_orders <-
cdrv_orders_tsv |>
  separate(doc_date, into = c("d1", "d2", "d3"), sep = "-", remove = F) |>
  mutate(date_precision = case_when(
    is.na(doc_date) | doc_date=="" ~ NA, 
    d2=="00" ~ "y",
    d3=="00" ~ "ym",
    .default = "ymd"
  )) |>
  # get rid of 00 before parsing date.
  # \\b here should cover years like 1700 if needed; but check before using
  mutate(doc_date = str_replace_all(doc_date, "\\b00", "01")) |>
  # parse date
  mutate(date = parse_date_time(doc_date, "ymd")) |>
  mutate(doc_year = year(date))  |>
  relocate(date, date_precision, doc_year, .after = doc_date) |>
  # why have i not done this...?
  arrange(date, order_id) |>
  select(-d1, -d2, -d3, -to_place, -to_area) |>
  # add name(s) of removed.
  left_join(
    cdrv_names |>
      mutate(glue_name = glue("{stdgiv} {surname}")) |>
      group_by(order_id) |>
      summarise(removed = glue_collapse(glue_name, sep = ", ", last = " and ")) |>
      ungroup(), by="order_id"
  ) |>
  mutate(vagrant = if_else(is.na(vagrant), "no", "yes")) |>
  mutate(has_exam = if_else(is.na(link_exam_id), "no", "yes")) 


cdrv_orders_geo <-
  cdrv_orders |>
  # geo data
  left_join(cdrv_geo_csv |>
              select(order_id, to_place, to_county, city, country, lat, lng, src), by="order_id") |>
  # leaflet doesn't like NAs! only drops about 30. 
  filter(!is.na(lat)) |>
  
  # display date, varying precision
  mutate(m = as.character(month(date, label=TRUE, abbr = FALSE))) |>
  mutate(nice_date = case_when(
    date_precision=="y" ~ as.character(doc_year),
    date_precision=="ym" ~ paste(m, doc_year ),
    date_precision=="ymd" ~ paste( day(date) , m, doc_year)
  )) |>
  relocate(nice_date, .after = doc_date) |>
  mutate(order_url = glue("{ll_url_base}{img}")) 

```




```{r}
ojs_define(ojs_cdrv_geo = cdrv_orders_geo)
```

```{ojs}
ojsCdrvGeo = transpose(ojs_cdrv_geo)
```


```{ojs} 
function createRVGeoData(data) {
  const geoData = {
    type: 'FeatureCollection',
    crs: {type: 'name', properties: {name: 'urn:ogc:def:crs:EPSG::27700'}},
    features: []
  };
  data.map(h => {
    geoData.features.push({
      type: 'Feature',
      properties: {
        id: h.order_id,
        place: h.to_place,
        county: h.to_county,
        removed: h.removed,
        vagrant: h.vagrant, // yes/no for display
        date: h.nice_date,
        year: h.doc_year,
        order_url: h.order_url,
        has_exam: h.has_exam // "yes"/"no"
      },
      geometry: {
        type: 'Point',
        coordinates: [h.lng, h.lat]
      }
    });
  });
  return geoData;
}
```


## The Map



```{ojs}

viewof filtersRV = Inputs.form({
// linked exams (most)
  hasExam: Inputs.checkbox(["yes", "no"], {label: "Linked exam(s)", value: ["yes", "no"]}) ,
  // a vagrant?
  isVagrant: Inputs.checkbox(["yes", "no"], {label: "Vagrant", value: ["yes", "no"]}) 
}
,
// https://observablehq.com/@tophtucker/use-inputs-form-with-a-template
// https://developer.mozilla.org/en-US/docs/Web/CSS/How_to/Layout_cookbook/Column_layouts
  {
    template: (inputs) => htl.html`<div style="display: flex; flex-wrap: wrap;" class="styled">
    <style>
  div.styled { border: 1px solid #d3d3d3; background-color: #f5f5f5; margin: 5px; padding: 5px; font-size: 80%; }
  div.styled form {border: 1px solid white;}
  div.styled form>label { width: 130px; padding:5px; }
</style>
  ${Object.values(inputs)}
</div>`
  }
)
  
```



```{ojs}

map = {
  // Make a container in the Observable DOM to hold the map
  let parent = DOM.element('div', { style: `width:${width}px; height:600px` });
  // Make sure Observable renders the parent container so Leaflet can figure out its size
  yield parent;
  
  let map = L.map(parent)
      .setView([54.3, -3.5], 6); // uk+ire // higher numbers are more zoomed in
  
  // any of no-referrer-when-downgrade, origin, origin-when-cross-origin, strict-origin, or strict-origin-when-cross-origin
  // osmMap = L.tileLayer(osmUrl, {attribution: osmAttrib, maxZoom: 19, referrerPolicy: 'strict-origin-when-cross-origin'});
  let osmTileLayer = L.tileLayer( 'http://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png', {
      attribution: '&copy; <a href="https://www.openstreetmap.org/copyright">OpenStreetMap</a>',
      maxZoom: 16,
      minZoom: 5,
      referrerPolicy: 'strict-origin-when-cross-origin'
  })
  .addTo(map);
  
/*
  const key = 'Qm3IHfglBpG81gp4ZK4G';
  let mtTileLayer = L.tileLayer(`https://api.maptiler.com/maps/base-v4/{z}/{x}/{y}.png?key=${key}`,{ //style URL
    tileSize: 512,
    zoomOffset: -1,
    minZoom: 4,
    attribution: "\u003ca href=\"https://www.maptiler.com/copyright/\" target=\"_blank\"\u003e\u0026copy; MapTiler\u003c/a\u003e \u003ca href=\"https://www.openstreetmap.org/copyright\" target=\"_blank\"\u003e\u0026copy; OpenStreetMap contributors\u003c/a\u003e",
    crossOrigin: true
}).addTo(map);
  */



  const geojsonData = createRVGeoData(ojsCdrvGeo);
  
  const mcg = markerCluster({
            spiderfyDistanceMultiplier: 1.8,
            maxClusterRadius: 50
            })
            .addTo(map);
	
	// single markers with a simple popup for testing. is L.geoJson the same as L.geoJSON ? seems to be
  let markers = L.geoJSON(null, {
  
  
    onEachFeature: function (feature, layer) {
    
      const data = feature.properties;
      const tip = `<div class="popup">${data.removed}: ${data.place} (${data.year})</div>`; // + `<br/>${data.place}, ${data.county} (${data.year})</div>`;
      
      layer.bindPopup(tip); // click.
      //layer.bindTooltip(tip, {sticky: true}); // hover. 
 
    },
    
    pointToLayer: function (feature, latlng) {
			return L.circleMarker(latlng, {
				radius: 6, 
				fillColor: "#CC6000", // a bit darker than "#ff7800" 
				color: "#000",
				weight: 1, // width of the circle border.
				opacity: 1,
				fillOpacity: 1
			});
     }
  }) ;  
  
// Once you have populated your Leaflet GeoJSON Layer Group (typically with geojsonLayer.addData(geoJsonObject), then instead of adding that group to your map, simply add it into your MarkerClusterGroup:
  markers.addData(geojsonData).addTo(mcg);

  // This is to make these variables accessible in other Observable cells.
  // It is not necessary for regular HTML web pages.
  
  // AHHHHHHH you need to take map out of here and use mcg *instead*
  mutable mapVariables = [mcg, markers];
}

```



```{ojs}

mapFilters = {
  // This handles the changes 
  // In normal HTML environment, this would be inside the slider event handling function.

  mapVariables[1].eachLayer(l => {
    if( 
        filtersRV["hasExam"].includes(l.feature.properties.has_exam) 
        && filtersRV["isVagrant"].includes(l.feature.properties.vagrant)
        )  
    {
      l.addTo(mapVariables[0]);
    } else {
      l.removeFrom(mapVariables[0]);
    }
  })

  return 'howdy'; // do i need this return line at all? probably not...
}
```

```{ojs}
mutable mapVariables = null;
```


```{ojs}
// import markercluster plugin
markerCluster = L, require('leaflet.markercluster@1.1.0').catch(() => L.markerClusterGroup)
```


```{ojs}
markerClusterCSS = html`<link href='${resolve('leaflet.markercluster@1.1.0/dist/MarkerCluster.Default.css')}' rel='stylesheet' />`
```


## OpenStreetMap problem

ongoing problem

BUT

- not on all browsers: 
  - problem on Firefox (and Waterfox)
  - Chrome and Safari seem OK

(also it doesn't happen on localhost, only online, which makes testing more of a pain)


[OSM Wiki](https://wiki.openstreetmap.org/wiki/Blocked_tiles)

> If the tile mentions "Referer is required" then you were automatically blocked because your application/website is not sending HTTP Referer headers, but doing so is required by the tile usage policy. This block will resolve itself once you make the necessary changes to your application, without further intervention necessary.

> For websites/web applications, this can be fixed by setting your Referrer-Policy header to any of no-referrer-when-downgrade, origin, origin-when-cross-origin, strict-origin, or strict-origin-when-cross-origin. 

but doesn't mention why it would vary between browsers...


OSM forum discussions focusing on Leaflet

- [possible solution but might not still be working](https://community.openstreetmap.org/t/access-blocked-on-osm-org/141419/27)
- [referer policy links](https://community.openstreetmap.org/t/access-blocked-leaflet-tiles-just-me/135657/14)
- [github example](https://github.com/openstreetmap/leaflet-osm/pull/51/changes)


## Resources

Leaflet:

- [Leaflet](https://leafletjs.com/) 
- [Leaflet.markercluster plugin](https://github.com/Leaflet/Leaflet.markercluster)

Observable:

- [Learning Leaflet](https://observablehq.com/@rahulbot/learning-leaflet) (Observable Notebook)
- [Leaflet with Online GeoJSON and Slider](https://observablehq.com/d/de61852a430b89e9) (Observable Notebook with filtering method)
- [Chicago Homicides Leaflet Map](https://observablehq.com/@randomfractals/chicago-homicides-heatmap) (Observable Notebook with markerclusters)

