---
title: "Visualising residuals with scatterplots"
date: "2025-09-25"
categories: 
  - correlation
  - statistics
execute: 
  warning: false
  message: false
---

## Notes

I've seen this type of scatterplot used to visualise residuals; came across [a tutorial for making them](https://drsimonj.svbtle.com/visualising-residuals).

It's also a bit of practice for statistical modelling using `lm()`.

Using [Playfair's Data on Wages and the Price of Wheat](https://friendly.github.io/HistData/reference/Wheat.html) from the HistData package. (lm() might not be the best model for this...)


```{r}
#| code-fold: true

library(scales) 
library(janitor) 
library(glue)
library(tidyverse)

library(patchwork)
library(ggthemes)
theme_set(theme_minimal())  

library(mindseyedata)
library(HistData)


playfair_wheat <-
Wheat |>
  clean_names("snake") |>
  filter(if_all(everything(), ~ !is.na(.x)))
```


## Examples

The data (after dropping a few NAs).

```{r}
head(playfair_wheat)
```

What does a simple scatterplot look like?

```{r}
playfair_wheat |>
  ggplot(aes(wages, wheat)) +
  geom_point()
```

Steps to make the residuals plot:

1. Fit a regression model to predict variable (Y).
2. Obtain the predicted and residual values associated with each observation on (Y).
3. Plot the actual and predicted values of (Y) so that they are distinguishable, but connected.
4. Use the residuals to make an aesthetic adjustment to highlight points which are poorly predicted by the model.


I'll treat wheat prices as the response variable (could easily be the other way round with this data).

```{r}
# lm(outcome/response_variable [y] ~ predictor/explanatory_variable [x], data=df)

fit <- lm(wheat ~ wages, data = playfair_wheat)  # Fit the model

# find a way to do this in a tidy pipeline...
playfair_wheat$predicted <- predict(fit)   # Save the predicted values
playfair_wheat$residuals <- residuals(fit) # Save the residual values

summary(fit) 
```
With the predicted/residual values added.

```{r}
head(playfair_wheat)
```

The tutorial shows various ways to adjust a scatterplot to highlight the residuals. 

Steps:

1. plot the actual data 
2. plot the predicted values in a way that makes them distinguishable from the actual values (eg by shape)
3. connect the actual data points with their corresponding predicted value using geom_segment():
4. make adjustments: eg 
  - fade out connection lines with alpha; 
  - add best fit line with geom_smooth()
5. final adjustments (to actual or predicted) to highlight the size of the residual. 

At step 4:

```{r}
playfair_wheat |>
  # set up with outcome variable on y-axis
ggplot(aes(x = wages, y = wheat)) +  
  # (3) connect actual and predicted; (4) use alpha to fade lines
  geom_segment(aes(xend = wages, yend = predicted), alpha = .2) +  
  # (1) plot the actual points
  geom_point() + 
  # (2) add the predicted values as open circles.
  geom_point(aes(y = predicted), shape=1) + 
  # (4) add the regression line
  geom_smooth(method = "lm", se = FALSE, color = "lightgrey") 
```

Step 5 examples

- change alpha, colour, size of actual values based on absolute value of residuals
- change colour according to whether difference is plus or minus and/or by how much.

(the example colours are terrible though, find somethng nicer)

```{r}
playfair_wheat |>
ggplot(aes(x = wages, y = wheat)) + 
  geom_segment(aes(xend = wages, yend = predicted), alpha = .2) + 
  geom_point(aes(colour=residuals, size=abs(residuals)), show.legend = F) + 
  geom_point(aes(y = predicted), shape = 1) + 
  geom_smooth(method = "lm", se = FALSE, color = "lightgrey") + 
  scale_color_gradient2(low = "blue", mid = "white", high = "red")  
```

The tutorial also shows how to do faceted versions to compare regressions on different variables if you have more than two. I'll need different data for that!
