---
title: "Waffle charts"
date: "2025-08-02"
categories: 
  - part-to-whole
execute: 
  warning: false
  message: false
---

## Notes

[Waffle charts](https://r-graph-gallery.com/waffle.html) (aka "square pie charts") represent categorical data via a grid of small squares forming a larger square or rectangle, which resembles a waffle. 

### strengths

- visually pleasing alternative to pie charts or stacked bar charts for showing how parts relate to the whole
- can be used for either amounts or proportions
- easier to see/compare smaller amounts or percentages than with a stacked bar

### limitations

- maybe less good for showing fine differences in proportional version (all % are rounded to whole number)

### variations

- "parliament charts" in various shapes


## Examples

Using cheshire_petitions in mindseyedata and [Apprenticeship Disputes in the Lord Mayor’s Court of London 1573-1723](https://doi.org/10.5255/UKDA-SN-8490-1)


```{r includes}
#| code-fold: true

library(scales) 
library(janitor) 
library(readxl)
library(glue)
library(tidyverse)

library(ggthemes)
theme_set(theme_minimal())  

library(waffle)

library(mindseyedata)

apprentices_xlsx <-
  read_excel(here::here("data/apprenticeship_disputes.xlsx"), 
             sheet = "Main index") |>
  clean_names("snake") 

apprentices <-
  apprentices_xlsx |>
  rename(year_binding=x21, year_petition=x27) |>
  # minor cleaning, some inconsistencies in data
  mutate(across(
    c(year_binding, year_petition, result_of_the_proceedings, reason_for_the_petition), 
                ~str_remove_all(., "x|z"))) |>
  mutate(across(
    c(year_binding, year_petition, result_of_the_proceedings, reason_for_the_petition), 
                ~na_if(., ""))) |>
  
  mutate(result_of_the_proceedings = str_to_upper(result_of_the_proceedings)) |>
  mutate(reason_for_the_petition = case_when(
    reason_for_the_petition %in% c("e", "p") ~ str_to_upper(reason_for_the_petition),
    .default = reason_for_the_petition
  )) |>
  mutate(across(c(year_binding, year_petition), parse_number))  |>
  mutate(reason = if_else(str_detect(reason_for_the_petition, "^U"), 
                          "U", 
                          reason_for_the_petition )) |>
  mutate(reason = case_match(
    reason,
    "B" ~ "Not bound correctly",
    "C" ~ "Unreasonable chastisement",
    "E" ~ "Non-enrollment",
    "F" ~ "Master a foreigner",
    "I" ~ "Instruction not provided",
    "N" ~ "Necessaries not provided",
    "P" ~ "Expelled from service",
    "S" ~ "Not in business in the city", 
    "U" ~ "Bound under the age of 14"
  ))  |>
  mutate(result = case_match(
    result_of_the_proceedings,
    "A" ~ "agreement reached",
    "D" ~ "apprentice discharged",
    "N" ~ "no decision",
    "W" ~ "jury decides for defendant"
  )) |>
  # remove 2 and 3 digit year of binding which are probably errors
  mutate(year_binding=if_else(year_binding<1500, NA, year_binding)) |>
  mutate(p_petition = case_when(
    is.na(year_petition) ~ NA,
    year_petition<1650 ~ "1569-1649",
    year_petition<1675 ~ "1650-1674",
    year_petition<1700 ~ "1675-1699",
    .default = "1700-1722"
  )) |>
  mutate(p_binding = case_when(
    is.na(year_binding) ~ NA,
    year_binding<1650 ~ "1573-1649",
    year_binding<1675 ~ "1650-1674",
    year_binding<1700 ~ "1675-1699",
    .default = "1700-1723"
  )) |>
  mutate(apprentice_years = if_else(
    str_detect(number_of_years_of_the_apprenticeship, "[0-9]"), 
    number_of_years_of_the_apprenticeship,
    "")) |>
  mutate(apprentice_years = parse_number(apprentice_years))  |>
  select(id, number_of_petition, p_binding, p_petition, reason, result, apprentice_years)  


cheshire_petitions <-
  cheshire_petitions |>
  select(petition_id, year, topic, petition_type, petition_gender, 
         response, petitioner, reference) |>
  mutate(petition_type = str_remove(petition_type, " *on behalf")) 

```


### waffle package

A popular and easy to use ggplot-based package is [waffle](https://github.com/hrbrmstr/waffle), by Bob Rudis.

For comparison though, let's start with a quick look at two more familiar types of chart that waffle charts are often used to replace, the stacked bar chart and pie chart (which is essentially a bar chart turned into a circle).

There is a difficulty with the apprentices data: one category (`non-enrollment`) is much larger than all the rest put together, which is not an uncommon experience with history data.

```{r}
apprentices |>
  count(reason) |>
  # remove a few very small categories and NAs
  filter(n>100) |>
  ggplot(aes("", n, fill=reason)) +
  geom_col(position = "fill") +
  scale_fill_ptol() +
  scale_y_continuous(labels = percent_format()) +
  labs(fill="complaint", y="% of cases", x=NULL, 
       title = "Apprentices' complaints")
```

The pie chart is prettier but has the same problem; adding percentage labels to the pie helps. I'd generally only use pie charts for a smaller number of categories. (Based on [code here](https://r-graph-gallery.com/piechart-ggplot2.html).)

```{r}
apprentices |>
  count(reason) |>
  filter(n>100) |>
  mutate(prop = round(n / sum(n)*100),1) |>
  mutate(prop_label = paste0(prop, "%")) |>
  ggplot(aes(x="", y=prop, fill=reason)) +
  geom_col(width = 1, color="white") +
  geom_text(aes(label = prop_label), position = position_stack(vjust=0.5), 
            color = "white", size=3) +
  coord_polar("y", start=0) +
  scale_fill_ptol() +
  theme_void()  +
  labs(title="Apprentices' complaints", fill="complaint")  
```


caveat: categories can get broken up over multiple rows, which might be less readable? Very tiny categories (less than 1%) are omitted. 


```{r}
apprentices |>
  count(reason) |>
  filter(n>100) |> 
  ggplot(aes(fill=reason, values=n)) +
  geom_waffle(colour="white", n_rows = 10, make_proportional = TRUE) +
  coord_equal() +
  scale_fill_ptol() +
  theme_enhance_waffle() +
  labs(title = "Apprentices' complaints", caption = "1 square=1% of cases", 
       fill="complaint")

```

a non-proportional waffle chart, with one square per case for the top 6 topics of Cheshire TPOP petitions.

```{r}
cheshire_petitions  |>
  count(topic) |>
  filter(n>25) |> 
  ggplot(aes(fill=topic, values=n)) +
  geom_waffle(colour="white", n_rows = 20) +
  coord_equal() +
  scale_fill_colorblind() +
  theme_enhance_waffle() +
  theme(axis.ticks = element_blank()) +
  labs(caption = "1 square=1 petition")
```


### add another dimension 

As the waffles are based on {ggplot}, it's straightforward to incorporate faceting (aka small multiples or trellis charts) to add another variable to make comparisons.

Here, the apprentices data is split up into four time periods.

```{r}
apprentices |>
  filter(!is.na(p_petition)) |>
  add_count(reason) |>
  filter(n>100) |> 
  count(p_petition, reason, name="nn") |>
  ggplot(aes(fill=reason, values=nn)) +
  geom_waffle(colour="white", n_rows = 10, make_proportional = TRUE) +
  coord_equal() +
  facet_wrap(~p_petition) +
  scale_fill_ptol() +
  theme_enhance_waffle() +
    theme(axis.ticks = element_blank()) +
  labs(caption = "1 square=1% of cases")  
```

Breakdown of Cheshire petitions by topic and petitioner type ("collective" petitions are from groups like "the inhabitants of Runcorn" rather than named individuals). 

```{r}
cheshire_petitions |>
  add_count(topic, name="n_topic") |>
  filter(n_topic>25) |> 
  count(petition_type, topic, n_topic) |>
  ggplot(aes(fill=petition_type, values=n)) +
  geom_waffle(colour="white", n_rows = 10, make_proportional = TRUE) +
  coord_equal() +
  facet_wrap(~topic) +
  scale_fill_colorblind() +
  theme_enhance_waffle() +
  theme(axis.ticks = element_blank()) +
  labs(caption = "1 square=1% of petitions", fill="petitioner\ntype")
```


### parliament charts

a semi-circular version of a waffle chart used to visualise party representation in legislatures. R package for making these (and other shapes), [ggparliament](https://zmeers.github.io/ggparliament/).

This one's the Australian House of Representatives in 2010. 

```{r}
library(ggparliament)
aus_rep <-
election_data |> 
  filter(country=="Australia" & house=="Representatives" & year==2010)

aus_house <- parliament_data(election_data = aus_rep,
                                       type = "horseshoe",
                                       parl_rows = 4,
                                       party_seats = aus_rep$seats)

aus_house |>
    ggplot(aes(x = x, y = y, colour = party_short)) +
    geom_parliament_seats() + 
    theme_ggparliament() +
    labs(colour = NULL, 
         title = "Australian Parliament 2010") +
    scale_colour_manual(values = aus_house$colour, 
                        limits = aus_house$party_short) 
```


```{r}
#| include: false

# colorspace::swatchplot(aus_rep$colour, cvd=TRUE)
```


## Resources

- [R Graph Gallery](https://r-graph-gallery.com/waffle.html)

R packages:

- B Rudis (2023). *waffle*. R package version 1.0.2. https://github.com/hrbrmstr/waffle
- R Hickman, Z Meers, T Leeper (2024). *ggparliament: Parliament Plots*. R package version 3.1.6. https://github.com/zmeers/ggparliament
