---
title: "automate frequency tables"
author: "Sharon Howard"
date: "2025-10-06"
categories: [variables, combinations, frequency tables, note]
execute:
  warning: false
  message: false
---

## Includes

```{r}
source(here::here('R/shared.R'))
```


## Notes

The scenario: I have a dataset with several variables and want to make frequency tables of all the 2-way combinations (maybe even 3-way combinations).


## Examples

Using Old Bailey trials data.

```{r}
trials_cat <-
obv_trials |>
  select(deft_speech_cat, deft_gender, offence_cat, verdict_cat) |>
  filter(!is.na(deft_gender), verdict_cat %in% c("guilty", "notGuilty")) 
```


```{r}
trials_cat |>
  head(10)
```


Given a tibble in the format above, with each variable in its own column,
it's amazingly easy to do 1-way frequency tables and get a (named) list of tables. 

```{r}
too_easy <-
  map(trials_cat, table)

too_easy
```


so then you can run, eg, chi-squared goodness of fit tests on all of them at once...

```{r}
map(too_easy, chisq.test)
```

2-way tables need a little bit more work.
Instead of just using a single tibble, you want a list of tibbles, each one containing a pair of variables.

[see post on automating counts for the gory details.]

```{r}
cat_names <- 
  trials_cat |>
  names() |> 
  set_names() 

cat_names_list <-
  combn(cat_names, m=2 ) |>  
  as_tibble(.name_repair = ~c("sg", "so", "sv", "go", "gv", "ov") ) |> 
  as.list()

all_select <- function(data, vars){
  data |> select(all_of(vars))
}

trial_cat_groups <-
cat_names_list |>
  map(~all_select(trials_cat, .x))
```


then it's just map() and table() again.

```{r}
(trial_cat_tabs <-
  map(trial_cat_groups, table)
 )
```

Run chi-squared test of independence on all the tables.

```{r}
map(trial_cat_tabs, chisq.test)
```




