---
title: "variable as column name"
author: "Sharon Howard"
date: "2025-09-30"
categories: [variables, naming things]
execute:
  warning: false
  message: false
---

## Includes

```{r}
source(here::here('R/shared.R'))
```

## Notes

I'm lazy. I don't like typing the same thing over and over. *Especially* when I got data from somewhere which has long column names.

So sometimes I want to do something like

```{r}
var <- "a_very_long_column_name_in_my_dataset"
```

and then reuse the stored short `var` instead of typing out a_very_long_column_name_in_my_dataset.

Examples below may also work in writing functions; compare with the various TILs on naming things/functions. 


## Examples

Using a sample taken from the [Addressing Health Pensions Database](https://beta.ukdataservice.ac.uk/datacatalogue/studies/study?id=9325#!/details)

```{r}
addressing_health <- 
  read_csv(here::here("data/addressing_health_sample1000.csv"))
```


```{r}
addressing_health |>
  names()
```

Let's say I want to mess around with the columns `length_of_service_years` and `ppg_allowance_decimal`.

```{r}
(losy <- "length_of_service_years")
(pad <- "ppg_allowance_decimal")
```



## in tidyverse pipes

### select

use `all_of()` or `any_of()` (seem to be interchangeable)

```{r}
addressing_health |> 
  select(any_of(pad)) |>
  head(10)
```

can select more than one with usual tidy-select stuff

```{r}
addressing_health |> 
  select(any_of(c(losy, pad))) |>
  head(10)
```



### group by

use `group_by_at()` 

```{r}
addressing_health |>
  group_by_at(losy) |>
  summarise(n = n()) |>
  head(10)
```

or `across()` + `all_of()` (or `any_of`) 

```{r}
addressing_health |>
  group_by(across(all_of(losy))) |>
  summarise(n = n()) |>
  head(10)
```




### filter 

use `get()`


```{r}
addressing_health |> 
  filter(get(losy) <= 8)
```

### mutate

on the *right hand side* of a mutate, use `get()`

```{r}
addressing_health |>
  select(all_of(c(losy, pad))) |>
  mutate(p = get(losy)/get(pad)) |>
  head(10)
```

On the *left hand side* of a mutate (eg, you want to base the new column name on the old one) is slightly more complex: use !! and :=

```{r}
addressing_health |>
  select(all_of(c(losy, pad))) |>
  filter(get(losy) > 45) |>
  # nb !! goes outside paste()
  mutate(!! paste0("p_", losy) := get(losy)/get(pad))
```


if for some reason you want to reuse the variable as a string on the righthand side (eg make a new grouping name?)

```{r}
addressing_health |>
  select(all_of(c(losy, pad))) |>
  mutate(new_grp = paste0(losy, "_2")) |>
  head(10)
```


you could also make the new variable name beforehand rather than doing paste inside mutate

```{r}
newvarname <- paste0("p_", losy)

addressing_health |> 
  select(all_of(c(losy, pad))) |>
  mutate( !! newvarname := get(losy)/get(pad) ) |>
  head(10)
```

### summarise 

same as mutate, I think.

```{r}
addressing_health |> 
  summarise(ml = mean(get(losy)), !! paste0("mean_", pad) := mean(get(pad)))
```



## working with vectors

Let's say I want to get the mean, median and standard deviation of each column, as separate vectors rather than in a dataframe.
 
This is largely a lesson in subsetting... but
NB that the shorthand form x$y doesn't work; you have to use x[[y]].

([Hadley Wickham explains](https://adv-r.hadley.nz/subsetting.html#section-1) this is because x$var gets translated as x[["var"]] instead of x[[var]])


```{r}
mean_losy <- mean(addressing_health[[losy]])
median_losy <- median(addressing_health[[losy]])
sd_losy <- sd(addressing_health[[losy]])

mean_losy
median_losy
sd_losy
```

```{r}
mean_pension <- mean(addressing_health[[pad]])
median_pension <- median(addressing_health[[pad]])
sd_pension <- sd(addressing_health[[pad]])

mean_pension
median_pension
sd_pension
```


Or I want to get the length of service columnn - don't need get() or anything else this way

```{r}
head( addressing_health[losy] )
```

get the first 3 rows of the pension allowance column

```{r}
addressing_health[1:3, pad] 
```



## fun with frequency tables

Using the historydata judges data for this.

```{r}
g <- "gender"
bps <- "birthplace_state"
```

```{r}
table(judges_people[[bps]], judges_people[[g]]) |>
  head(10)
```

or, to make the table using `with()`, you want `get()` again. I imagine something similar applies if you want to use `xtab()`.

```{r}
with(judges_people, table(get(bps), get(g))) |>
  head(10)
```


