---
title: "pretty() breaks"
date: "2025-09-09"
categories: 
  - base
  - scales
  - ggplot
execute: 
  warning: false
  message: false
---

## Notes

- pretty() is a base R function to "compute a sequence of about n+1 equally spaced ‘round’ values which cover the range of the values in x". 
- handy to tweak breaks inside ggplot scale_*_ functions if you don't like the ggplot defaults, without having to work anything out for yourself

[helpful SO comment](https://stackoverflow.com/a/62496287/7281022) (my emphasis)

> the way pretty works is not easily predictable. I usually remind myself that it was implemented to create **reasonable** tick marks for charts. In a chart you usually need *a good first and last value and equal steps in between*...


```{r}
#| code-fold: true

library(scales) 
library(janitor) 
library(glue)
library(tidyverse)

library(ggthemes)
theme_set(theme_minimal())  

library(ggridges)

library(mindseyedata)
library(historydata)

```


## Examples

```{r}
# count of topic for the scale
petitions_topic <-
  cheshire_petitions |> 
  count(topic)

(pet_pretty_break <-
  pretty(petitions_topic$n, n=10))
```


```{r}
cheshire_petitions |>
  mutate(petition_type = str_remove(petition_type, " *on behalf")) |>
  count(topic, petition_type) |>
  mutate(topic = fct_rev(topic)) |>
  ggplot(aes(n, topic, fill = petition_type)) +
  geom_col() +
  scale_x_continuous(breaks = pet_pretty_break) +
  scale_fill_tableau()
```

```{r}
methodists |>
  ggplot(aes(year, members_total)) +
  geom_point() +
  scale_y_continuous(breaks = pretty(methodists$members_total, n=10)) +
  scale_x_continuous(breaks = pretty(methodists$year, n=10))
```

Uses outside ggplot? 



## More

The scales package has some fancier [axis breaks functions](https://scales.r-lib.org/reference/index.html#axis-breaks) including eg for log scales and date/time scales. To try out some time!


## Resources

[cran](https://search.r-project.org/R/refmans/base/html/pretty.html)