---
title: "frequency tables - multidimensional (base R)"
author: "Sharon Howard"
date: "2025-09-25"
categories: [frequency tables, base]
execute:
  warning: false
  message: false
---


## Includes

```{r}
source(here::here('R/shared.R'))

petitions <-
  cheshire_petitions |>
  mutate(petition_type = str_remove(petition_type, " *on behalf"))
```


## Notes

- see also vcd::structable which does essentially the same thing as ftable (but a base R method may be useful if you don't want vcd)


## Examples

if you give `table()` three variables it will make a list of 2-way tables, breaking down by the last variable.

```{r}
(petitions_table <- 
   table(petitions$topic, petitions$petition_gender, petitions$petition_type) )
```

`xtabs` does the same thing

```{r}
xtabs(~ topic + petition_gender + petition_type, data=petitions) 
```



`ftable()` makes a flattened 3-way table

you can do it in one step (play around with order of variables...)

```{r}
ftable(petition_type ~ topic + petition_gender, petitions) 
```

or you can give it your list of 2-way tables and it will flatten them into a 3-way. (vcd::structable will do that as well)

```{r}
ftable(petitions_table)
```












