---
title: "fill in missing values with complete()"
author: "Sharon Howard"
date: "2025-06-30"
categories: [combinations, NAs]
slug: til-complete-counted-combinations-20250630
execute:
  warning: false
  message: false
---

## Includes

```{r}
source(here::here('R/shared.R'))
```

## Notes

-  it's not exactly missing data, but sometimes you want something like this for ggplot 
-  needs a pair of columns 
-  fill=list(n=0) will fill missings with 0 instead of NA 






## Examples

A data frame of counts. Most judges won't appear in all court types, but I need explicit 0s for those cases.

```{r}
judges100_appts |> 
  count(judge_id, court_type) |> 
  complete(judge_id, court_type, fill=list(n=0)) 
```



A short list of biographical categories, as a vector. I want all the pairs of categories.

```{r}
(categories <- c("height", "weight", "age", "gender"))
```




It'll include identical pairs, which I probably won't want; I can filter those out. [See TILs on combn, expand_grid, for alternative approaches.]

```{r}
tibble(a=categories, b=categories) |>
  complete(a, b) |>
  filter(a != b) 
```


## Resources

* [https://tidyr.tidyverse.org/reference/complete.html](https://tidyr.tidyverse.org/reference/complete.html)

```{r}
#| include: false
# beyond_notability/bn_wikibase/themes/ppa/bn_work_queries.Rmd:161:  
 

```
