---
title: "subsetting in base R - notes and gotchas"
author: "Sharon Howard"
date: "2025-06-30"
categories: [wip, needs data, subset, base]
slug: til-subsetting-101-gotchas-to-understand-20250630
execute:
  warning: false
  message: false
---

## Includes

```{r}
source(here::here('R/shared.R'))
```

## Notes

-  penguins will work but would prefer history data! 
-  > The [ operator always returns an object of the same class as the original. It can be used to select multiple elements of an object 
-  > The [[ operator is used to extract elements of a list or a data frame. It can only be used to extract a single element and the class of the returned object will not necessarily be a list or data frame. 
-  > The $ operator is used to extract elements of a list or data frame by literal name. Its semantics are similar to that of [[. 
 
-  Q: commas. what is the , doing in this example [src?]?? removing it doens't seem to make any difference 
-  mtcars[, c("mpg", "cyl", "wt")]  
-  mtcars[c("mpg", "cyl", "wt")] 
 




## Example

```{r}
penguins[["bill_len"]] # bill_len col as a vector of values. 
penguins$bill_len # same again  
penguins[[3]]  # same again 
 
penguins[,3]  # depends...! this is interesting. with a DF I got a vector of values again, but after changing to a tibble it's a tibble with single column.  
penguins[3] # df/tibble with single col  
penguins[3,] # a df with all columns for the third row 
penguins[3,3] # third col in third row, as a vector or single cell tibble 
 
names(penguins[1]) # [1] "species" - name of the first col  
names(penguins["species"]) # [1] "species" [this seems a very silly thing to do...?] 
# names(penguins[species]) # error object species not found 
# names(penguins[[species]]) # ditto 
names(penguins[["species"]]) # NULL 
 
 

```


## Resources

* [https://adv-r.hadley.nz/subsetting.html](https://adv-r.hadley.nz/subsetting.html)

```{r}
#| include: false
# atb/atb/_stuff/bor_biographies_sample_2204.Rmd:291:  
# small_bills/sb_explore/themes/sb_counties.Rmd:304:  
 

```
