---
title: "yaml doesn't have to be at the start of a rmd document! "
author: "Sharon Howard"
date: "2025-06-30"
categories: [nocode, yaml]
slug: til-nocode-yaml-anywhere-20250630
execute:
  warning: false
  message: false
---

## Includes

```{r}
source(here::here('R/shared.R'))
```

## Notes

-  You can use inline R code (see Section 3.1) anywhere in an Rmd document, including the YAML metadata section. This means some YAML metadata can be dynamically generated with inline R code, such as the document title. 
-  If your title depends on an R variable created later in the document, you may add the title field in a later YAML section,  
-  The title works in this case because Pandoc can read any number of YAML sections in a document (and merge them). 
 
 
 







## Resources

* [https://bookdown.org/yihui/rmarkdown-cookbook/dynamic-yaml.html](https://bookdown.org/yihui/rmarkdown-cookbook/dynamic-yaml.html)

```{r}
#| include: false
# atb/atb/documentation/earlier-experimental-might-still-be-useful/atb_att_values-to-pdf.Rmd:146:  
 

```
