---
title: "function to make alt text for dataviz using glue() "
author: "Sharon Howard"
date: "2025-06-30"
categories: [strings, glue, accessibility]
slug: til-functions-glue-alt-text-20250630
execute:
  warning: false
  message: false
---

## Includes

```{r}
source(here::here('R/shared.R'))
```

## Notes

-  formula "chart type of type of data where reason for including chart" 
-  what was the source for this? bluesky/twitter? is it a quote? looks like one 
-  > Playing around with @AmyCesal boiler plate example for writing alt text, but adapted to a #rstats function."chart type of type of data where reason for including chart" along with the data source. 
-  but how is this added to output ??? 


## Function

```{r} 
write_alt_text <- function( 
  chart_type,  
  type_of_data,  
  reason,  
  source 
){ 
  glue::glue( 
    "{chart_type} of {type_of_data} where {reason}. \n\nData source from {source}" 
    ) 
} 

```




## Resources

* [https://medium.com/nightingale/writing-alt-text-for-data-visualization-2a218ef43f81](https://medium.com/nightingale/writing-alt-text-for-data-visualization-2a218ef43f81)

```{r}
#| include: false
# r_reference/dataviz/accessibility.Rmd:18:  
 

```
