---
title: "de-select multiple columns with all_of() or any_of() "
author: "Sharon Howard"
date: "2025-06-30"
categories: [select]
slug: til-select-not-all-any-of-20250630
execute:
  warning: false
  message: false
---

## Includes

```{r}
source(here::here('R/shared.R'))
```

## Notes

-  neater than using minus sign? though less concise. and perhaps more usable inside functions? 
-  [got this from Kieran Healy, but i'm not sure where] 
-  > all_of() is for strict selection. If any of the variables in the character vector is missing, an error is thrown. 
-  > any_of() doesn't check for missing variables. It is especially useful with negative selections, when you would like to make sure a variable is removed. 
 
 




## Example

```{r}
cheshire_petitions |> 
 select(!all_of(c("petition_gender", "year"))) 

```


## Resources

* [https://tidyselect.r-lib.org/reference/all_of.html](https://tidyselect.r-lib.org/reference/all_of.html)

```{r}
#| include: false
# beyond_notability/bn_wikibase/snippets.Rmd:20:  
# beyond_notability/bn_wikibase/themes/bn_more_networks.Rmd:326:  
 

```
