---
title: "pdf options"
author: "Sharon Howard"
date: "2025-06-30"
categories: [nocode, export, pdf]
slug: til-nocode-export-pdf-options-20250630
execute:
  warning: false
  message: false
---

## Includes

```{r}
source(here::here('R/shared.R'))
```

## Notes

-  NB making nice PDFs will drive you round the bend 
-  NB that {tinytex} was installed to handle latex dependencies (might need reinstalling) 
 
-  pdf options in yaml header 
-  under pdf_document: 
-    df_print: not sure if this is doing anything much for this stuff since you're using tables... 
-    latex_engine: important to specify this as xelatex or lualatex (idk what diffs) to have a lot of the other options, esp font stuff 
 
-  top level options: 
-  mainfont: can use any system font; use quotes if more than one word 
-  monofont: if you need it 
-  linestretch: increases line spacing but *not* padding inside tables 
-  documentclass: (eg) scrreprt  need to find out more about these 
-  geometry: several page settings - paper size, orientation, margins 
-    margin=xxcm for all margins; to set them separately: left=xx, right=xx, top=xx, bottom=xx 
 
-  getting more advanced... header-includes and \usepackage - tinytex should hopefully handle anything not already installed. 
-  header-includes: 
-     - \usepackage[default]{sourcesanspro} 
 








```{r}
#| include: false
# atb/atb/documentation/earlier-experimental-might-still-be-useful/more-tables.Rmd:25:  

```
